---
title: "Speech Intelligibility Project"
author: "Sophie Shan, Esther Zhou, Wanbing Wang, Dan Huang"
format: pdf
editor: visual
---

# Section 1: Read in Data and Load Packages

First, we load the necessary packages for reading in the data and then load the dataset. We expect there to be 115 children.

## Load necessary packages

```{r}
#| echo: true
#| message: false
#| warning: false
#| label: load-packages

# --- Data I/O & Wrangling ---
library(readxl)      # read Excel files
library(tidyverse)   # data wrangling and visualization (dplyr, ggplot2, tidyr, purrr)

# --- Reporting & Tables ---
library(knitr)       # render output tables and format reports
library(table1)      # formatted descriptive (Table 1) summaries

# --- Statistical Modeling ---
library(tidymodels)  # model utilities (broom for tidy coefficient output)
library(lmerTest)    # linear mixed models with Satterthwaite p-values
library(emmeans)     # estimated marginal means and pairwise contrasts
library(broom.mixed) # tidy() method for lme4/lmerTest mixed-effects model objects

# --- Statistical Testing & Diagnostics ---
library(irr)         # inter-rater reliability (ICC)
library(ggpubr)      # normality diagnostic plots (density, Q-Q)
library(TOSTER)      # equivalence testing (TOST, bootstrapped)
```

## Read in Data

```{r}
#| label: read-in-data

speech <- read_excel("/Users/sophieshan/Desktop/Group Project/1-Data/SWI_Master_Results.xlsx",
                    sheet = "Sheet1")
head(speech)
dim(speech)
```

# Section 2: Prep Data for Analysis

## Relabel Column Names

```{r}
#| label: relabel-columns

colnames(speech)[3] <- "Cleft_Type"
colnames(speech)[26] <- "st_fric_affr"
colnames(speech)[29] <- "VP_Status"
```

## Make Variables into Factors

```{r}
#| label: make-factors

#rename study cohorts 
speech$Cohort <- factor(
  speech$Cohort,
  levels = c(1, 2, 3, 4),
  labels = c("CLP", "CP", "TD", "OM")
)

#rename sex 
speech$Sex <- factor(
  speech$Sex,
  levels = c(1, 2),
  labels = c("Male", "Female")
)

#rename VP status
speech$VP_Status <- factor(speech$VP_Status,
                           levels = c(1, 2),
                           labels = c("Adequate", "Inadequate"))
```

## Create a Cohort x Sex Variable

```{r}
#| label: cohort-sex

speech$GroupSex <- factor(
  paste(speech$Cohort, speech$Sex),
  levels = c(
    "CLP Male", "CLP Female",
    "CP Male",  "CP Female",
    "TD Male",  "TD Female",
    "OM Male",  "OM Female"
  )
)
```

## Add Listener Group for ID 30310

```{r}
#| label: listener-group-30310

#check to see which listeners are in for 30310
speech %>%
    filter(ID == "30310") %>%
    dplyr::select(L1, L2, L3)

#check to see where listener 75, 76, 74 are
speech %>%
    filter(L1 == 75 |  L1 == 76 | L1 == 74) %>%
    dplyr::select(ID, Cohort, Cleft_Type, Sex, Age, L1, L2, L3, L_Group)

#based on this, we see that the most likely listener group for ID 30310 is Group 12
speech[speech$ID == 30310, "L_Group"] <- 12

#make L_Group into a factor
speech <- speech %>%
    mutate(L_Group = as.factor(L_Group))
```

## Create Long Version

```{r}
#| label: create-long

#Extract variables for simple analysis
speech_subset <- speech %>%
    dplyr::select(ID, Cohort, Cleft_Type, Sex, Age, 
                  VP_Status, L1_SWI, L2_SWI, L3_SWI, L_Group)

#Transform Data to long format
speech_subset_long <- speech_subset %>%
      pivot_longer(                                                                                                
          cols = c(L1_SWI, L2_SWI, L3_SWI),           
          names_to = "Listener",                                                                                             values_to = "SWI")    

#Give unique names to each listener within group (there should be 66 unique listeners)
speech_subset_long <- speech_subset_long %>%                                                                       
      mutate(Listener = paste0("G", L_Group, "_", str_extract(Listener, "L\\d+"))) %>%
      mutate(Listener = as.factor(Listener)) %>%
      mutate(L_Group = as.factor(L_Group))
dim(speech_subset_long)

#double check
n_distinct(speech_subset_long$Listener)  

#Remove the row with one NA observation (L3 for ID 30310)
speech_subset_long <- speech_subset_long %>%
    filter(!(is.na(SWI == T)))
dim(speech_subset_long)

#summary
summary(speech_subset_long)
```

# Section 3: Figures and Tables

## Figure 1a

Distribution Speech of Mean Intelligibility Scores by Group and Sex Mean Speech Intelligibility vs. Age by Group. Boxplots show median, IQR, and range by group and sex.

```{r}
#| label: boxplot-mswi-group-sex-4groups

# Custom x-axis positions: pairs are (Male, Female) for each group (CLP, CP, TD, OM)
# Gaps between pairs create visual separation between groups
positions4 <- c(1, 2.2,  4.8, 6.0,  8.6, 9.8,  12.4, 13.6)

# Draw boxplot with boxes placed at custom positions instead of default 1:8
# col alternates light/dark gray for Male/Female within each group
bp <- boxplot(M_SWI ~ GroupSex, data = speech,
              at = positions4,         # place boxes at custom x positions
              xaxt = "n",              # suppress default x-axis (added manually below)
              xlab = "",
              ylab = "Mean Speech Intelligibility Score",
              cex.main = 0.9,
              col = rep(c("gray85", "gray60"), 4),  # light = Male, dark = Female
              medlwd = 3,              # thicker median line
              whisklty = 2,            # dashed whiskers
              boxwex = 0.9,            # box width
              xlim = c(0, 15)
)

# Add Male/Female labels below each box
text(
  x = positions4,
  y = par("usr")[3] - 3.5,            # slightly below the plot area
  labels = rep(c("Male", "Female"), 4),
  xpd = TRUE                          # allow drawing outside plot region
)

# Add group labels (CLP, CP, TD, OM) centered between each Male/Female pair
text(
  x = c(mean(c(1, 2.2)),
        mean(c(4.8, 6.0)),
        mean(c(8.6, 9.8)),
        mean(c(12.4, 13.6))),
  y = par("usr")[3] - 8,              # further below than the Male/Female labels
  labels = c("CLP", "CP", "TD", "OM"),
  xpd = TRUE
)

# Add x-axis title
mtext("Group and Sex", side = 1, line = 5)

# Overlay median values as text just above each median line (row 3 of bp$stats)
text(positions4, bp$stats[3, ] + 1,
     round(bp$stats[3, ], 1),
     cex = 0.7, font = 2)
```

## Figure 1b

Distribution Speech of Mean Intelligibility Scores by Group and Sex Mean Speech Intelligibility vs. Age by Group. Boxplots show median, IQR, and range by group and sex.

```{r}
#| label: boxplot-mswi-group-sex-3groups

# Custom x-axis positions: pairs are (Male, Female) for each group (CLP, CP, Control)
# Gaps between pairs create visual separation between groups
positions3 <- c(1, 2.2, 4.8, 6.0, 8.6, 9.8)

# Merge TD and OM into a single Control group, then create a 3-level GroupSex variable
speech_3group <- speech %>%
    mutate(Cohort3 = case_when(
        Cohort == "CLP"            ~ "CLP",
        Cohort == "CP"             ~ "CP",
        Cohort %in% c("TD", "OM") ~ "Control"
    )) %>%
    mutate(GroupSex3 = factor(
        paste(Cohort3, Sex),
        levels = c("CLP Male", "CLP Female",
                   "CP Male",  "CP Female",
                   "Control Male", "Control Female")
    ))

# Draw boxplot with boxes placed at custom positions instead of default 1:6
# col alternates light/dark gray for Male/Female within each group
bp3 <- boxplot(M_SWI ~ GroupSex3, data = speech_3group,
               at = positions3,         # place boxes at custom x positions
               xaxt = "n",              # suppress default x-axis (added manually below)
               xlab = "",
  ylab = "Mean Speech Intelligibility Score",
               cex.main = 0.9,
               col = rep(c("gray85", "gray60"), 3),  # light = Male, dark = Female
               medlwd = 3,              # thicker median line
               whisklty = 2,            # dashed whiskers
               boxwex = 0.9,            # box width
  xlim = c(0, 11)
)

# Add Male/Female labels below each box
text(
  x = positions3,
  y = par("usr")[3] - 3.5,             # slightly below the plot area
  labels = rep(c("Male", "Female"), 3),
  xpd = TRUE                           # allow drawing outside plot region
)

# Add group labels (CLP, CP, Control) centered between each Male/Female pair
text(
  x = c(mean(c(1, 2.2)),
        mean(c(4.8, 6.0)),
        mean(c(8.6, 9.8))),
  y = par("usr")[3] - 8,               # further below than the Male/Female labels
  labels = c("CLP", "CP", "Control"),
  xpd = TRUE
)

# Add x-axis title
mtext("Group and Sex", side = 1, line = 5)

# Overlay median values as text just above each median line (row 3 of bp3$stats)
text(positions3, bp3$stats[3, ] + 1,
     round(bp3$stats[3, ], 1),
     cex = 0.7, font = 2)
```

## Figure 2

Scatter plot with points colored by group, points represent individuals. Optional trend lines may be added.

```{r}
#| label: scatter-plot

ggplot(speech, aes(x = Age, y = M_SWI, color = Cohort, shape = Cohort)) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Age",
    y = "Mean Speech Intelligibility Score",
    color = "Group",
    shape = "Group"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

## Figure 3

Boxplot split by mean intelligence scores by VP status.

```{r}
#| label: boxplot-VP-status

boxplot(M_SWI ~ VP_Status,
        data = speech,
        xlab = "VP Status",
        ylab = "Mean Speech Intelligibility Score",
        col = c("lightblue", "lightpink"))
```

# Section 4: Preliminary Analysis

## ICC

To quantify observer agreement before running the main model. If the ICC is very high ($>0.90$), averaging across observers and running a simpler ANOVA/linear regression okay.

Assume that for each listener, the same 3 listeners rated each of the study subjects. Note that in the raw data, groups 1,2,3,6,9,10,11,12,13,17,22 all had more than 3 listener IDs per listener group.

```{r}
#| label: icc

# Split data by listener group so ICC can be computed separately for each group
speech_L_Group <- speech %>%
    dplyr::select(L_Group, L1_SWI, L2_SWI, L3_SWI) %>%
    group_split(L_Group)

# Function to compute two-way agreement ICC for a single listener group
# Columns 2:4 are the three listener SWI scores; unit = "single" gives ICC for one rater
icc_per_group <- function(group_df){
    icc_val <- icc(group_df[, 2:4], model = "twoway",
                   type = "agreement", unit = "single")
    return(cbind(group_df$L_Group[1], icc_val$value))
}

# Apply function across all groups and combine into a single data frame
# V1 = group ID, V2 = ICC value
icc_results <- lapply(speech_L_Group, icc_per_group)
icc_results_df <- as.data.frame(do.call(rbind, icc_results))
icc_results_df$V2 <- round(icc_results_df$V2, 3)
icc_results_df

# Proportion of groups with poor agreement (ICC < 0.50)
prop.table(table(icc_results_df$V2 < 0.50))

# Distribution of ICC values across all listener groups
hist(icc_results_df$V2,
     main = "Distribution of ICC Values Across Listener Groups",
     xlab = "ICC (Two-Way Agreement, Single Rater)",
     ylab = "Number of Listener Groups")
```

## Comparing OM to TD

A preliminary analysis was conducted to assess whether the TD and OM groups were sufficiently similar in mean SWI scores to justify pooling. Normality was first assessed using the Shapiro-Wilk test. According to Shapiro-Wilk normality test, the p-value \> 0.05 implying that the distribution of the data are not significantly different from normal distribution. We cannot assume normality.

```{r}
#| label: om-td-normality

#TD
TD <- speech %>%
    filter(Cohort == "TD")
ggdensity(TD$M_SWI,
          main = "Density plot of mean SWI",
          xlab = "Mean SWI")
shapiro.test(TD$M_SWI)

#OM
OM <- speech %>%
    filter(Cohort == "OM")
ggdensity(OM$M_SWI,
          main = "Density plot of mean SWI",
          xlab = "Mean SWI")
shapiro.test(OM$M_SWI)
```

A bootstrapped equivalence test (TOST) was then conducted, as this approach allows us to actively test for similarity rather than simply failing to detect a difference. Bootstrapped methods were used given that normality assumptions were not met. An equivalence margin of $\pm5$ points was specified, representing a difference deemed negligible on the 0--100 SWI scale.

```{r}
#| label: om-td-tests

# Bootstrapped TOST: tests whether the mean difference falls within [-5, +5]
# hypothesis = "EQU": equivalence hypothesis (both bounds must be satisfied)
# paired = FALSE: independent samples (TD and OM are separate groups)
# eqb = 5: equivalence margin of ±5 points on the 0-100 SWI scale
# alpha = 0.05: significance level for each one-sided test
# boot_ci = "stud": studentized bootstrap confidence intervals 
# (more accurate than percentile CI)

set.seed(1)

boot_t_TOST(x = TD$M_SWI, y = OM$M_SWI,
            hypothesis = "EQU",
            paired = FALSE,
            eqb = 5,
            alpha = 0.05,
            boot_ci = c("stud"))
```

The observed mean difference between TD and OM was 3.08 points (90% bootstrapped CI \[0.87 5.68\]). Formal equivalence within ±5 points could not be confirmed as the upper bound was not satisfied ($p = .101$), however the difference was considered clinically negligible (\~3% of the full scale range). The two groups were therefore combined into a single control group for all subsequent analyses.

```{r}
#| label: combine-OM-TD

#for wide dataset
speech_analysis <- speech %>%
    mutate(new_Cohort = factor(
        if_else(Cohort == "TD" | Cohort == "OM", "Control", as.character(Cohort)),
        levels = c("CLP", "CP", "Control")  # CLP as reference group
    ))
table(speech_analysis$Cohort, speech_analysis$new_Cohort)

#for long dataset
speech_long_analysis <- speech_subset_long %>%
    mutate(new_Cohort = factor(
        if_else(Cohort == "TD" | Cohort == "OM", "Control", as.character(Cohort)),
        levels = c("CLP", "CP", "Control")  # CLP as reference group
    ))
table(speech_long_analysis$Cohort, speech_long_analysis$new_Cohort)
```

## Examining the Influence of Inadequate VP Status

Now, we examine the influence of children with inadequate velopharyngeal (VP) status. We want to see if we should include them in the primary/secondary analysis (this was brought up as question by the investigator).

```{r}
#| label: vp-descriptive

# Descriptive summary of mean SWI by VP status, cohort, and sex
# Used to quantify the disproportionate influence of 
#inadequate VP children on CLP female estimates
speech_analysis %>%
    group_by(VP_Status, new_Cohort, Sex) %>%
  summarise(
    mean_SWI = round(mean(M_SWI), 1),
    sd       = round(sd(M_SWI), 2),
    n        = n(),
    .groups  = "drop"
  )
```

Children with inadequate VP were almost entirely concentrated in the CLP group and had substantially lower intelligibility scores than their adequate-VP peers. We see that the female with inadequate-VP status had nearly 36 points lower than adequate-VP CLP females. Inadequate-VP status and CLP is almost 1:1 (near-complete confounding). We decide to remove.

```{r}
#for wide dataset
speech_analysis_no_vp <- speech_analysis %>%
    filter(VP_Status == "Adequate")
table(speech_analysis_no_vp$VP_Status)

#for long dataset 
speech_long_analysis_no_vp <- speech_long_analysis %>%
    filter(VP_Status == "Adequate")
table(speech_long_analysis_no_vp$VP_Status)
```

# Section 5: Aim 1

## Table 1

```{r}
#| label: table-1

table1(~ Cleft_Type + Sex + Age + VP_Status + M_SWI| new_Cohort, 
       data = speech_analysis_no_vp)  
```

## Primary Analysis

We use LRT-based forward selection (`add1()`) to determine which pairwise interactions to include in the primary model. None of the three candidate interactions were significant: `new_Cohort:Age` (p=0.677), `new_Cohort:Sex` (p=0.517), and `Age:Sex` (p=0.398). The final primary model retains main effects only.

```{r}
#| label: lmm-interaction-LRT

# Step 1: Screen all pairwise interactions
# Fit main-effects-only base model with ML (REML cannot be used for LRT on fixed effects)
speech_lmm_base_ml <- lmer(SWI ~ new_Cohort + Age + Sex + 
                             (1 | ID),
                           REML = FALSE,
                           data = speech_long_analysis_no_vp)

# add1 adds each candidate interaction to the base model one at a time and tests
# improvement in fit via LRT — equivalent to forward selection for interactions
# Results: new_Cohort:Age (p=0.677), new_Cohort:Sex (p=0.517), Age:Sex (p=0.398)
# None are significant — retain main effects only
add1(speech_lmm_base_ml,
     scope = ~ (new_Cohort + Age + Sex)^2,
     test = "Chisq")
```

Based on the LRT interaction screening above, the primary model does not need any interaction terms. Now we check which random effects structure is appropriate.

```{r}
#| label: lmm-re-check

# Each child contributes 3 listener scores, so a child-level random intercept
# is needed to account for within-child correlation across listener ratings.
# Fit with both L_Group and ID to compare their variance contributions.
speech_lmm_re_check <- lmer(SWI ~ new_Cohort + Age + Sex +
                               (1 | L_Group) + (1 | ID),
                             REML = TRUE,
                             data = speech_long_analysis_no_vp)

# L_Group variance collapses to exactly 0 (singular fit) once child-level ID is included,
# confirming listener group adds nothing — simplify to (1 | ID) only
VarCorr(speech_lmm_re_check)
```

The child-level random intercept (SD ≈ 5.97) accounts for ~82% of total variance, reflecting that each child has a strong personal baseline intelligibility. L_Group variance collapses to exactly 0 (singular fit) once child-level clustering is included, so the random effects structure is simplified to `(1 | ID)`.

```{r}
#| label: lmm-final

speech_lmm_final <- lmer(SWI ~ new_Cohort + Age + Sex +
                           (1 | ID),
                         REML = TRUE,
                         data = speech_long_analysis_no_vp)

summary(speech_lmm_final)
```

Let's compute the estimated marginal means (EMMs) for each cohort.

```{r}
#| label: lmm-estimated-marginal-means

# EMMs are model-adjusted means that account for all other covariates (Age, Sex)

emm <- emmeans(speech_lmm_final, ~ new_Cohort)
emm   
```

Next, we look at the pairwise contrasts: Control vs CLP and CP vs CLP.

```{r}
#| label: lmm-pairwise-contrasts

# ref = 1 sets CLP as the reference group (first level of new_Cohort)
# Hochberg adjustment controls familywise error rate across the two comparisons
emm_contrast <- contrast(emm, method = "trt.vs.ctrl", ref = 1, adjust = "hochberg")

# infer = TRUE returns estimates, 95% CIs, and p-values together
summary(emm_contrast, infer = TRUE)  
```

Now, we take a look at the linear mixed model diagnostics.

```{r}
#| label: lmm-model-diagnostics

par(mfrow = c(2, 2))

# Residuals vs Fitted
plot(fitted(speech_lmm_final), residuals(speech_lmm_final),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# Normal Q-Q
qqnorm(residuals(speech_lmm_final), main = "Normal Q-Q")
qqline(residuals(speech_lmm_final), col = "red")

# Residual Histogram
hist(residuals(speech_lmm_final),
     xlab = "Residuals", ylab = "Frequency",
     main = "Residual Histogram",
     col = "grey")

# Scale-Location
plot(fitted(speech_lmm_final), sqrt(abs(residuals(speech_lmm_final))),
     xlab = "Fitted values", ylab = "|Scaled residuals|",
     main = "Scale-Location")

par(mfrow = c(1, 1))
```

## Sensitivity Analysis

### Linear Regression with Mean SWI

This sensitivity analysis uses the child-level mean SWI (averaged across the 3 listeners) as the outcome instead of individual listener scores. Because each child contributes one observation, no random effect is needed. We fit the simplified model without L_Group and use an F-test to confirm it adds no explanatory value.

```{r}
#| label: ols-mean-SWI-no-vp

# Primary OLS model: cohort, age, and sex — mirrors the fixed-effects structure of the LMM
speech_ols_no_vp <- lm(M_SWI ~ new_Cohort + Age + Sex,
                   data = speech_analysis_no_vp)

summary(speech_ols_no_vp)

# Augmented model with L_Group as a fixed effect to test whether listener group
# explains additional variance in mean SWI beyond cohort, age, and sex
speech_ols_no_vp_lg <- lm(M_SWI ~ new_Cohort + Age + Sex + as.factor(L_Group),
                   data = speech_analysis_no_vp)

# F-test: L_Group adds no significant variance (p = 0.532) — consistent with
# the LMM finding that listener group variance collapses to near zero once
# child-level clustering is accounted for
anova(speech_ols_no_vp, speech_ols_no_vp_lg)
```

```{r}
#| label: ols-mean-SWI-no-vp-emm

# EMMs and pairwise contrasts from the simplified model, Hochberg-adjusted
emm_ols_no_vp <- emmeans(speech_ols_no_vp, ~ new_Cohort)

emm_ols_no_vp_contrast <- contrast(emm_ols_no_vp, method = "trt.vs.ctrl", 
                                   ref = 1, adjust = "hochberg")
summary(emm_ols_no_vp_contrast, infer = TRUE)  
```

### Including Inadequate VP Status

In the primary analysis, we excluded those 7 individuals with inadequate VP status. We now redo the analysis with them included. We include VP_Status as a main effect in the model but don't consider interaction effects with VP_Status because of singularity.

#### LMM

```{r}
#| label: sensitivity-include-vp
speech_lmm_vp <- lmer(SWI ~ new_Cohort + Age + Sex + VP_Status +
                        (1 | ID),
                   REML = TRUE,
                   data = speech_long_analysis)

summary(speech_lmm_vp)
```

We compute estimated marginal means.

```{r}
#| label: lmm-include-vp-estimated-marginal-means

# EMMs adjusted for Age, Sex, and VP_Status
emm_vp <- emmeans(speech_lmm_vp, ~ new_Cohort)
emm_vp   
```

Now, we compute pairwise contrasts using the same approach as the primary analysis, with Hochberg adjustment. The pairwise contrasts are extremely different from the primary analysis.

```{r}
#| label: lmm-include-vp-pairwise-contrasts

# Control vs CLP and CP vs CLP, Hochberg-adjusted
emm_vp_contrast <- contrast(emm_vp, method = "trt.vs.ctrl", 
                            ref = 1, adjust = "hochberg")

# infer = TRUE returns estimates, 95% CIs, and p-values together
summary(emm_vp_contrast, infer = TRUE)  
```

Now, let's take a look at the residuals.

```{r}
#| label: sensitivity-include-vp-residuals

par(mfrow = c(2, 2))

# Residuals vs Fitted
plot(fitted(speech_lmm_vp), residuals(speech_lmm_vp),
     xlab = "Fitted values", ylab = "Residuals",
     main = "Residuals vs Fitted")
abline(h = 0, col = "red", lty = 2)

# Normal Q-Q
qqnorm(residuals(speech_lmm_vp), main = "Normal Q-Q")
qqline(residuals(speech_lmm_vp), col = "red")

# Residual Histogram
hist(residuals(speech_lmm_vp),
     xlab = "Residuals", ylab = "Frequency",
     main = "Residual Histogram",
     col = "grey")

# Scale-Location
plot(fitted(speech_lmm_vp), sqrt(abs(residuals(speech_lmm_vp))),
     xlab = "Fitted values", ylab = "|Scaled residuals|",
     main = "Scale-Location")

par(mfrow = c(1, 1))
```

#### Linear Regression

We refit the OLS model on the full dataset with VP_Status included as a covariate, and again confirm that L_Group adds no explanatory value.

```{r}
#| label: ols-mean-SWI

# OLS on child-level mean SWI with VP_Status as additional covariate
speech_ols <- lm(M_SWI ~ new_Cohort + Age + Sex + VP_Status,
                   data = speech_analysis)

summary(speech_ols)

# Augmented model with L_Group to verify it adds no variance in the full dataset
speech_ols_lg <- lm(M_SWI ~ new_Cohort + Age + Sex + VP_Status + as.factor(L_Group),
                    data = speech_analysis)

# F-test: confirms L_Group is not needed even when inadequate VP children are included
anova(speech_ols, speech_ols_lg)
```

Now, we compute the marginal means and pairwise contrasts.

```{r}
#| label: ols-mean-SWI-emm

# EMMs and pairwise contrasts from the simplified model, Hochberg-adjusted
emm_ols <- emmeans(speech_ols, ~ new_Cohort)

emm_ols_contrast <- contrast(emm_ols, method = "trt.vs.ctrl", ref = 1, 
                             adjust = "hochberg")
summary(emm_ols_contrast, infer = TRUE)  
```

# Section 6: Aim 2

For Aim 2, we want to see if there are differences between the 11 phonetic contrasts for the three groups.

```{r}
#| label: table-2

table1(~ alv_pal + alv_velar + cl_sing + con_null + fr_bk +
       fric_affr + hi_lo + r_w + st_fric_affr + st_nas + vd_vl                                                             | new_Cohort, data = speech_analysis_no_vp)
```

## Primary Analysis

We define the 11 phonetic contrasts of interest and their corresponding human-readable labels used throughout the analysis.

```{r}
#| label: aim2-setup

# Column names in the dataset corresponding to each phonetic contrast
contrasts_vars <- c("alv_pal", "alv_velar", "cl_sing", "con_null",
                    "fr_bk", "fric_affr", "hi_lo", "r_w",
                    "st_fric_affr", "st_nas", "vd_vl")

# Human-readable labels for each contrast (used in output tables)
contrast_labels <- c("Alveolar-palatal sibilant (e.g., see–she)",
                     "Alveolar-velar stop (e.g., tap–cap)",
                     "Cluster-singleton (e.g., stop–top)",
                     "Initial consonant-null (e.g., pat–at)",
                     "Front-back vowel (e.g., fill–full)",
                     "Fricative-affricate (e.g., ship–chip)",
                     "High-low vowel (e.g., beat–bit)",
                     "Liquid-glide (e.g., ray–way)",
                     "Stop-fricative/affricate (e.g., base–vase)",
                     "Stop-nasal (e.g., bop–mop)",
                     "Voiced-voiceless stop (e.g., big–pig)")
```

A separate linear model is fit for each phonetic contrast, regressing the contrast score on cohort, sex, and age. Cohort EMMs are then extracted from each model and stacked into a single data frame for downstream testing.

```{r}
#| label: aim2-fit-models

# For each phonetic contrast outcome, fit a linear model with cohort, sex, and age
# No cohort:sex interaction — consistent with Aim 1 interaction screening
fit_contrast_model <- function(outcome_var, data) {
  formula <- as.formula(
    paste(outcome_var, "~ new_Cohort + Sex + Age")
  )
  lm(formula, data = data)
}

# Fit one model per contrast and name the list for downstream extraction
models <- map(contrasts_vars, fit_contrast_model, data = speech_analysis_no_vp)
names(models) <- contrasts_vars

# Extract cohort contrasts using emmeans, averaged over Sex
extract_coefs <- function(model, contrast_label) {
  emm <- emmeans(model, ~ new_Cohort)
  contrast(emm, method = "trt.vs.ctrl", ref = 1) %>%
    summary(infer = TRUE) %>%
    as_tibble() %>%
    mutate(contrast_var = contrast_label) %>%
    rename(conf.low = lower.CL, conf.high = upper.CL) %>%
    select(contrast_var, contrast, estimate, conf.low, conf.high, p.value)
}

# Stack contrasts from all models into one data frame
results_raw <- map2_dfr(models, contrast_labels, extract_coefs)
```

We apply a two-family multiple testing procedure with Hochberg correction. Family 1 tests CLP vs. Control for all 11 contrasts. Only contrasts significant in Family 1 are carried forward to Family 2, which tests CLP vs. CP. This sequential gating controls the familywise error rate across both comparisons.

```{r}
#| label: aim2-multiple-comparisons

# Family 1: CLP vs. Control across all 11 contrasts
# Hochberg correction applied across the 11 tests
family1 <- results_raw %>%
  filter(contrast == "Control - CLP") %>%
  arrange(p.value) %>%
  mutate(
    p_hochberg_f1 = p.adjust(p.value, method = "hochberg"),
    reject_f1     = p_hochberg_f1 < 0.05
  )

# Family 2: CLP vs. CP, only for contrasts significant in Family 1
rejected_contrasts <- family1 %>%
  filter(reject_f1) %>%
  select(contrast_var)

family2 <- results_raw %>%
  filter(contrast == "CP - CLP") %>%
  inner_join(rejected_contrasts, by = "contrast_var") %>%
  arrange(p.value) %>%
  mutate(
    p_hochberg_f2 = p.adjust(p.value, method = "hochberg"),
    reject_f2     = p_hochberg_f2 < 0.05
  )
```

Family 1 and Family 2 results are joined and formatted into a summary table displaying the estimated group difference, 95% CI, and Hochberg-adjusted p-value for each phonetic contrast. Contrasts not carried into Family 2 are indicated with an em dash.

```{r}
#| label: aim2-results-table

# Helper to format estimate and 95% CI as a single string
fmt_ci <- function(est, lo, hi) {
  sprintf("%.1f (%.1f, %.1f)", est, lo, hi)
}

# Join Family 1 and Family 2 results and format for display
table3 <- family1 %>%
  select(contrast_var,
         est_control = estimate,
         lo_control  = conf.low,
         hi_control  = conf.high,
         p_adj_f1    = p_hochberg_f1) %>%
  left_join(
    family2 %>%
      select(contrast_var,
             est_cp   = estimate,
             lo_cp    = conf.low,
             hi_cp    = conf.high,
             p_adj_f2 = p_hochberg_f2),
    by = "contrast_var"
  ) %>%
  arrange(match(contrast_var, contrast_labels)) %>%
  mutate(
    `CLP vs. Control: β (95% CI)` = fmt_ci(est_control, lo_control, hi_control),
    `Adjusted p (Family 1)`       = ifelse(p_adj_f1 < 0.001,
                                           "<0.001",
                                           sprintf("%.3f", p_adj_f1)),
    `CLP vs. CP: β (95% CI)`      = ifelse(
      !is.na(est_cp),
      fmt_ci(est_cp, lo_cp, hi_cp),
      "—"
    ),
    # Contrasts not carried into Family 2 are marked with an em dash
    `Adjusted p (Family 2)`       = case_when(
      is.na(p_adj_f2)   ~ "—",
      p_adj_f2 < 0.001  ~ "<0.001",
      TRUE              ~ sprintf("%.3f", p_adj_f2)
    )
  ) %>%
  select(
    `Phonetic Contrast`           = contrast_var,
    `CLP vs. Control: β (95% CI)`,
    `Adjusted p (Family 1)`,
    `CLP vs. CP: β (95% CI)`,
    `Adjusted p (Family 2)`
  )

#display
table3 %>%                                                                                                          
    kable(digits = 3, caption = "Phonetic Contrast Results")
```
